\bf Causal Inference in Population Trends: Searching for Demographic Anomalies in Big Data

### 2020-01-24 MANUSCRIPT/manuscript.tex
All #'s should now be calculated in-text.

\if0\blind
{
  \title{\bf Causal Inference in Population Trends: A Technique for finding
Demographic Anomalies in Big Data}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Causal Inference in Population Trends: A Technique for finding
Demographic Anomalies in Big Data}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The proliferation of big data, wider access to advanced computing
platforms, and the development of powerful statistical algorithms can
uncover hidden anomalies in demographic data that have previously been
dismissed as noise. Here, we demonstrate how demographers can use causal
inference techniques and abductive reasoning to identify fertility and
mortality anamolies in state-level data trends by investigating
demographic time series data since 1999. In demonstrating our technique,
we uncover real spikes and dips in fertility and mortality over time,
distinguished from regular trend variations. We show that 38 states
exhibited at least one mortality anomaly over time, totalling more than
318,000 additional deaths, while an additional six states exhibited
periods of decline resuling in 164,000 fewer deaths. Twelve states
exhibited real, marked periods of declines in fertility, totalling more
than 240,000 fewer births than expected, and eleven states had periods
of excess births, resulting in an addition 134,000 births. Our results
show how researchers can detect demographic anomalies in trend data and
use those anomalies as a beginning point for further investigation into
the causes.
\end{abstract}

@@

\caption{\label{tab:unnamed-chunk-1}\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component. \label{sumtable}}
\centering
\begin{tabular}[t]{lrllrrr}
\toprule
Component & Anomalies & $y$ & $\hat{y}$ & $y-\hat{y}$ & $|y-\hat{y}|$ & \% of Total\\
\midrule
Fertility & 22 & 25.48M & 25.68M & -202,217 & 226,476 & 0.889\\
Mortality & 156 & 43.57M & 43.52M & 48,535 & 627,364 & 1.440\\
\bottomrule
\end{tabular}
\end{table}

\textbf{\autoref{sumtable}} report the overall number of anomalies we
detect of each type for births and deaths and some summary statistics
across all anomaly types. We find considerably more mortality anomalies
(n = 156) than fertility anomalies (n = 22). Given that anomalies are
always the produce of events \citep{Song2018ACM}, the finding of more
mortality than fertility anomalies is not surprising. Mortality is
likely to spike in response to a catastrophic event (like an earthquake
or terrorist attack) while the effects of a catastrophic event on
fertility is less predictable. One reason for this is that fertility is
linked to human decision-making more directly than mortality
\citep{stein2014couples} which results in more varied outcomes, so, for
example, researchers find that catastrophic weather events can increase
childbearing among those who already have a child, but not among the
childless \citep{Evans2008Hurricanebirth}. Some events, like the 1995
Oklahoma City bombings, resulted in both fertility and mortality
changes. However, changes in fertility manifest over a longer time
horizon after an event, and the relationship between a
fertility-inducing event and behavior change is less strong than the
relationship between a mortality-inducing event (such as a terrorist
incident) and death \citep{Rodgers2005OKBombing}. Of course, not all
events that impact fertility and mortality are catastrophic. Researchers
have documented that more commonplace events such as massive layoff
\citep{Venkataramani2019} or policy changes
\citep{Livingston2017Cannabis} can also create anomalies in mortality
patterns, but we have not found such evidence for fertility.

Consistent with more anomalies across the time lines, we find more
anomalous deaths (627,364) than anomalous births (226,476). Fertility
anomalies overwhelmingly tend toward lower fertility, with only a few
fertility anomalies yielding more births. This is consistent with
research findings that link high severity weather events to
significantly lower fertility \citep{Evans2008Hurricanebirth}.
Conversely, mortality anomalies tend to be more evenly split between
positive and negative anomalies, but tend toward more deaths rather than
less deaths. Again, this is not surprising, as the kinds of events that
increase death (e.g., plant closings are associated with increased
opioid death \citep{Venkataramani2019}) are more common than the events
that decrease death (e.g., cannabis legalization is associated with
decreased opioid deaths \citep{Livingston2017Cannabis}) These anomalous
deaths and births account for 1.44\% and 0.889\% over the entire time
series, respectively.

\textbf{\autoref{fig:summap}} maps these results. Three states exhibited
neither mortality nor fertility anomalies: Alaska, North Dakota, and
South Dakota. Another nine states exhibited just a single anomaly (DE,
DC, ID, IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts
exhibited the most anomalies with nine.

@@

Since the purpose of our paper is to demonstrate how demographers can
make use of causal inference technique, we will not discuss all of our
findings in this paper. Instead, we will use a few cases for
illustration. For mortality, we will use New York State as an example
(\textbf{\autoref{fig:mortnewhamp}a}). We identify seven anomalies in
the mortality time series for New York, all with t-statistics in excess
of 3.91, making these anomalies unlikely to be due to chance. The
algorithm correctly identifies September 2001 as an additive outlier
(2001:09 t=6.40) where mortality in that month was 1,628 higher than
anticipated. This mortality event is likely caused by the September 11
tragedy and the detection of this mortality event provides confidence in
our detection of other anomalies.

In \textbf{\autoref{fig:mortnewhamp}a}, notice the strong level shift
(LS) that occurs in February 2004 (2004:02 t=5.59) which prevented
almost 870 deaths per month. This shift totals more than 144,000 averted
deaths compared to the counter-factual time series and is the single
largest mortality protective anomaly among all states. This translates
to 4.5\% fewer deaths than expected over the time period. What is
driving this mortality protection? What policies did NY put into place
that might have contributed to this considerable mortality reduction?
What environmental conditions may have changed? These are the kinds of
questions that arise from our analyses. The purpose of our paper is not
to answer these questions, but our findings underscore the need for more
research that takes an abductive approach. By identifying anomalies
through an inductive process, researchers can then look for underlying
causes. Once those causes are identified, researchers can then use a
deductive process to see if such events predict other (or future)
anomalies.

To see the potential for combinging abductive reasoning with causal
inference, contrast the mortality protection in New York with the
enhanced mortality in New Hampshire
(\textbf{\autoref{fig:mortnewhamp}b}). In New Hampshire we detect two
significant level shifts (LS) in the monthly mortality data, first in
April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68).
These anomalies suggest New Hampshire experienced 9,700 more deaths
(+14\% more than expected) in a seven-year period beginning in early
2010. These are events not experienced by neighborhing states during the
same time period, and this is the single largest percentage mortality
increase/decrease we detected among all states. Not coincidentally, NH
has the second highest opioid-related mortality in the US
\citep{beetham2019access} and it is likely that we detect this epidemic
in our results. Isolating these anomalies and testing them against
opioid sales data might yeild intriguing results.

@@

To illustrate how researchers can examine fertility anomalies, we begin
with a clear temporary change (TC) outlier for fertility in Louisiana
(\textbf{\autoref{fig:fertla}a}). Here we detect two TC outliers, nearly
back-to-back in 2005 (2005:08 \emph{t} = -6; and 2005:10 \emph{t} = -5),
coinciding with the destruction of Hurricane Katrina from the same year.
The migration associated with Hurricane Katrina has garnered most of the
attention of demographers
\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
but the hurricane had a very clear impact on fertility behaviors too,
creating a mini ``baby bust'' in Louisiana with the departure of many
people in their childbearing years. We estimate 4,411 fewer births in
Louisiana compared to the counterfactual, likely attributable to the
Hurricane.

We also highlight a level shift (LS) outlier for fertility in
Connecticut. Here we detect a strong (\emph{t}= -3.71; -285
births/month) level shift toward lower fertility beginning in August
2009 (2009:08) that continues through the end of the period
(\textbf{\autoref{fig:fertla}b}). This LS toward lower fertility reduced
the \# of births by -28,752 or -8.57\% lower than the counterfactual
time series. The stock market crash of 2008, where the Dow Jones
Industrial Average had the largest single-day loss up to that point,
occured just 11 months before we detect a shift toward lower fertility.
It is possible that the trend toward lower fertility in August 2009 and
the stock market crash in September 2008 are linked, although, as stated
previously, we are speculating only to illustrate how causal inference
could work in an abductive research agenda.

@@

In \textbf{\autoref{fig:ferthawaii}a} we identify a single additive
anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with
a \emph{t}-statistic of 4.96, 14\% above the counter-factual time
series. This single, anomalous month is also the second highest monthly
births in the time series. We have no plausible explanation for this
anomaly. We do not believe this is simply a data error as the other
extreme values, September 2008 and February 2005 with the highest and
lowest recorded fertility respectively, were not identified as anomalous
events. Even if we were to assume this anomaly resulted from data entry
error, it remains an \emph{unaltered} data error in the Hawaiian monthly
fertility data.

This is contrast to \textbf{\autoref{fig:ferthawaii}b}, where we
identify a strange mortality reduction in Ohio (\emph{t}-stat: -4.27).
This LS is more than 1,100 deaths per month less than the counterfactual
time series, suggesting had nearly 40,000 fewer deaths since February
2015 than expected. This is the single-largest LS among all states. We
could not identify the potential policies Ohio might have put into place
to provide such a strong mortality protection.

>>>>

\if0\blind
{
  \title{\bf Causal Inference in Population Trends: Searching for Demographic
Anomalies in Big Data}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Causal Inference in Population Trends: Searching for Demographic
Anomalies in Big Data}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The proliferation of big data, wider access to advanced computing
platforms, and the development of powerful statistical algorithms can
uncover hidden anomalies in demographic data that have previously been
dismissed as noise. Here, we demonstrate how demographers can use causal
inference techniques and abductive reasoning to identify fertility and
mortality anamolies in state-level data trends by investigating
demographic time series data since 1999. In demonstrating our technique,
we uncover real spikes and dips in fertility and mortality over time,
distinguished from regular trend variations. We show that 36 states
exhibited at least one mortality anomaly over time, totalling nearly
300,000 additional deaths, while an additional seven states exhibited
periods of decline resuling in 250,000 fewer deaths. Ten states
exhibited real, marked periods of declines in fertility, totalling more
than 214,000 fewer births than expected, and 8 states had periods of
excess births, resulting in an addition 12,000 births. Our results show
how researchers can detect demographic anomalies in trend data and use
those anomalies as a beginning point for further investigation into the
causes.
\end{abstract}

@@

\caption{\label{tab:unnamed-chunk-2}\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component. \label{sumtable}}
\centering
\begin{tabular}[t]{lrllrrr}
\toprule
Component & Anomalies & $y$ & $\hat{y}$ & $y-\hat{y}$ & $|y-\hat{y}|$ & \% of Total\\
\midrule
Fertility & 22 & 25.48M & 25.68M & -202,217 & 226,476 & 0.889\\
Mortality & 156 & 43.57M & 43.52M & 48,535 & 627,364 & 1.440\\
\bottomrule
\end{tabular}
\end{table}

\textbf{\autoref{sumtable}} report the overall number of anomalies we
detect of each type for births and deaths and some summary statistics
across all anomaly types and \textbf{\autoref{fig:summap}} maps these
results. We find considerably more mortality anomalies (n = 156) than
fertility anomalies (n = 22). Given that anomalies are always the
product of events \citep{Song2018ACM}, finding more mortality than
fertility anomalies is not surprising. Mortality is likely to spike in
response to a catastrophic event (like an earthquake or terrorist
attack) or due to a disease outbreak while the effects of a catastrophic
event on fertility is less predictable. One reason for this is that
fertility is linked to human decision-making more directly than
mortality \citep{stein2014couples} which results in more varied
outcomes, so, for example, researchers find that catastrophic weather
events can increase childbearing among those who already have a child,
but not among the childless \citep{Evans2008Hurricanebirth}. Some
events, like the 1995 Oklahoma City bombings, resulted in both fertility
and mortality changes. However, changes in fertility manifest over a
longer time horizon after an event, and the relationship between a
fertility-inducing event and behavior change is less strong than the
relationship between a mortality-inducing event (such as a terrorist
incident) and death \citep{Rodgers2005OKBombing}. Of course, not all
events that impact fertility and mortality are catastrophic. Researchers
have documented that more commonplace events such as massive layoff
\citep{Venkataramani2019} or policy changes
\citep{Livingston2017Cannabis} can also create anomalies in mortality
patterns, but we have not found such evidence for fertility.

Consistent with more anomalies across the time series, we find more
anomalous deaths (627,364) than anomalous births (226,476). Fertility
anomalies overwhelmingly tend toward lower fertility, with only a few
fertility anomalies yielding more births. This is consistent with
research findings that link high severity weather events to
significantly lower fertility \citep{Evans2008Hurricanebirth}.
Conversely, mortality anomalies tend to be more evenly split between
positive and negative anomalies, but tend toward more deaths rather than
less deaths. Again, this is not surprising, as the kinds of events that
increase death (e.g., plant closings are associated with increased
opioid death \citep{Venkataramani2019}) are more common than the events
that decrease death (e.g., cannabis legalization is associated with
decreased opioid deaths \citep{Livingston2017Cannabis}). These anomalous
deaths and births account for 1.44\% and 0.889\% over the entire time
series, respectively.

As \textbf{\autoref{fig:summap}} shows, three states exhibited neither
mortality nor fertility anomalies: Alaska, North Dakota, and South
Dakota. Another nine states exhibited just a single anomaly (DE, DC, ID,
IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts exhibited
the most anomalies with nine.

@@

Since the purpose of our paper is to demonstrate how demographers can
make use of causal inference technique, we will not discuss all of our
findings here, in this paper. Instead, we will use a few cases for
illustration. For mortality, we will use New York State as an example
(\textbf{\autoref{fig:mortnewhamp}a}). We identify seven anomalies in
the mortality time series for New York, all with \emph{t}-statistics in
excess of 3.91, making these anomalies unlikely to be due to chance. The
algorithm correctly identifies September 2001 as an additive outlier
(2001:09 \emph{t}= 6.396) where there were 1,628 more deaths in that
month than anticipated. This mortality event is likely caused by the
September 11 tragedy and the detection of this mortality event provides
confidence in our detection of other anomalies.

In \textbf{\autoref{fig:mortnewhamp}a}, notice the strong level shift
(LS) that occurs in February 2004 (2004:02 \emph{t}= -5.594) which
prevented 869 deaths per month. This shift totals more than 144,000
averted deaths compared to the counter-factual time series and is the
single largest mortality protective anomaly among all states. This
translates to 6.90\% fewer deaths than expected over the time period.
What is driving this mortality protection? What policies did NY put into
place that might have contributed to this considerable mortality
reduction? What environmental conditions may have changed? These are the
kinds of questions that arise from our analyses. The purpose of our
paper is not to answer these questions, but our findings underscore the
need for more research that takes an abductive approach. By identifying
anomalies through an inductive process, researchers can then look for
underlying causes. Once those causes are identified, researchers can
then use a deductive process to see if such events predict other (or
future) anomalies.

To see the potential for combining abductive reasoning with causal
inference, contrast the mortality protection in New York with the
enhanced mortality in New Hampshire
(\textbf{\autoref{fig:mortnewhamp}b}). In New Hampshire we detect two
significant level shifts (LS) in the monthly mortality data, first in
April 2010 and again in November 2014 (2010:04 \emph{t}= 3.51; 2014:11
\emph{t}= 6.68). These anomalies suggest New Hampshire experienced 8,159
more deaths (+9.24\% more than expected) in a seven-year period
beginning in early 2010. These are events not experienced by neighboring
states during the same time period, and this is the single largest
percentage mortality increase/decrease we detected among all states. Not
coincidentally, NH has the second highest opioid-related mortality in
the US \citep{beetham2019access} and it is likely that we detect this
epidemic in our results. Isolating these anomalies and testing them
against opioid sales data might yield intriguing results.

@@

To illustrate how researchers can examine fertility anomalies, we begin
with a clear temporary change (TC) outlier for fertility in Louisiana
(\textbf{\autoref{fig:fertla}a}). Here we detect two TC outliers, nearly
back-to-back in 2005 (2005:08 \emph{t} = -6.002; and 2005:10 \emph{t} =
-4.997), coinciding with the destruction of Hurricane Katrina from the
same year. The migration associated with Hurricane Katrina has garnered
most of the attention of demographers
\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
but the hurricane had a very clear impact on fertility behaviors too,
creating a mini ``baby bust'' in Louisiana with the departure of many
people in their childbearing years. We estimate 4,411 fewer births in
Louisiana compared to the counterfactual, likely attributable to the
Hurricane.

We also highlight a level shift (LS) outlier for fertility in
Connecticut. Here we detect a strong (\emph{t}= -3.71 for -285
births/month) level shift toward lower fertility beginning in August
2009 (2009:08) that continues through the end of the period
(\textbf{\autoref{fig:fertla}b}). This LS toward lower fertility reduced
the \# of births by -28,752 or 8.57\%) lower than the counterfactual
time series. The stock market crash of 2008, where the Dow Jones
Industrial Average had the largest single-day loss up to that point,
occurred just 11 months before we detect a shift toward lower fertility.
It is possible that the trend toward lower fertility in August 2009 and
the stock market crash in September 2008 are linked, although, as stated
previously, we are speculating only to illustrate how causal inference
could work in an abductive research agenda.

@@

In \textbf{\autoref{fig:ferthawaii}a} we identify a single additive
anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with
a \emph{t}-statistic of 4.96, -14.9\% above the counter-factual time
series. This single, anomalous month is also the second highest monthly
births in the time series. We have no plausible explanation for this
anomaly. We do not believe this is simply a data error as the other
extreme values, September 2008 and February 2005 with the highest and
lowest recorded fertility respectively, were not identified as anomalous
events. Even if we were to assume this anomaly resulted from data entry
error, it remains an \emph{unaltered} data error in the Hawaiian monthly
fertility data.

This is contrast to \textbf{\autoref{fig:ferthawaii}b}, where we
identify a strange mortality reduction in Ohio (\emph{t}-stat: -4.27).
This LS is more than 1,157 deaths per month less than the counterfactual
time series, suggesting had nearly 40,000 fewer deaths since February
2015 than expected. This is the single-largest LS among all states. We
could not identify the potential policies Ohio might have put into place
to provide such a strong mortality protection.
